---
title: "WildFire_Project Analysis"
author: "Arundhati G., Zhongyang H., Adarsh K., Sumanth N."
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('ezids')
```


```{r }
WildfireData <- read.csv('final_wildfire.csv')
summary_nature=read.csv('summary_nature.csv')
summary_peoplecaused=read.csv('summary_peoplecaused.csv')
```

```{r}

Avg_Temp <- WildfireData$tair_day_livneh_vic
Avg_SoilMoisture <- WildfireData$soilmoist1_day_livneh_vic
Avg_Rainfall <- WildfireData$rainfall_day_livneh_vic

```

##Five Point Summary 
```{r, results="asis"}
xkablesummary(WildfireData)
```

## HISTOGRAMS

```{r, results="markup"}

library(ggplot2)

#Average Temperature
TempHist <- ggplot(WildfireData, aes(Avg_Temp))+ 
  geom_histogram(binwidth = 0.5, bins = 100, col="black", fill="light blue 2") +
  labs(x="Avg. Temp (C)", y="Frequency", title="HISTOGRAM: Average Temprature") 


#Average Soil Moisture
SoilHist <- ggplot(WildfireData, aes(Avg_SoilMoisture))+ 
  geom_histogram(binwidth = 0.5, bins = 100, col="black", fill="orange red 2") +
  labs(x="Avg. Soil Moisture", y="Frequency", title="HISTOGRAM: Average Soil Moisture") 

#Average Rainfall
RainHist <- ggplot(WildfireData, aes(Avg_Rainfall))+ 
  geom_histogram(binwidth = 0.5, bins = 100, col="black", fill="green 3") +
  labs(x="Avg. Rainfall", y="Frequency", title="HISTOGRAM: Average Rainfall") 

#Wildfire Count by Year
CountHist <- ggplot(WildfireData, aes(Year))+ 
  geom_histogram(binwidth = 0.10, bins = 100, col="black", fill="yellow", stat="count") +
  labs(x="Years", y="Frequency of Wildfires", title="Wildfires count by year", )

Histograms <- grid.arrange(TempHist, SoilHist, RainHist, CountHist, ncol=2, nrow=2)

ggsave("Histograms.jpg", plot = Histograms)
```


##Bar Graphs
```{r, results="markup"}

#Fire Size
FireBar <- ggplot(data = WildfireData, aes(x = FIRE_SIZE_CLASS)) +
  geom_bar(col="black", fill="orange")+
  labs(x="Fire Size Class", y="Frequency", title="Frequency of Wildfires by Size Classes") 


#Years
YearsBar <- ggplot(data = WildfireData, aes(x = Year)) +
  geom_bar(col="black", fill="yellow")+
  labs(x="Years", y="Frequency", title="Frequency of Wildfires by Year")

#Budget
BudgetBar <- ggplot(data = WildfireData, aes(x = Budget)) +
  geom_bar(col="black", fill="Pink 2")+
  labs(x="Budget", y="Frequency", title="Frequency of Wildfires by Budget")


grid.arrange(FireBar, YearsBar, nrow=2)


```

##Pie Charts
```{r, results="markup"}
lbls <- c("A", "B", "C", "D", "E", "F", "G", "E")
jpeg("sizeclasspie.jpeg")
sizeclasspie<- pie((table(WildfireData$FIRE_SIZE_CLASS)), col=rainbow(length(lbls)), main="Pie Chart of Fire Size Class")

```

```{r, results="markup"}

jpeg("cause_descrpPie.jpeg")

lbls <- c("Lightning", "Eqipment Use", "Smoking", "Campfire", "Debris Burning", "Railroad", "Arson", "Children", "Misc." )
cause_descrpPie <- pie((table(WildfireData$STAT_CAUSE_DESCR)), col=rainbow(length(lbls)), main="Pie Chart of Wildfire Cause")

```


## R Markdown
Hu Zhongyang part
intialize data

```{r }
final_fire=read.csv('final_wildfire.csv')
summary_nature=read.csv('summary_nature.csv')
summary_peoplecaused=read.csv('summary_peoplecaused.csv')
colnames(summary_nature)[4]='temperature'
colnames(summary_nature)[5]='soilmoisture'
colnames(summary_nature)[6]='rainfall'
colnames(summary_nature)[7]='nfire'
colnames(summary_peoplecaused)[4]='temperature'
colnames(summary_peoplecaused)[5]='soilmoisture'
colnames(summary_peoplecaused)[6]='rainfall'
colnames(summary_peoplecaused)[7]='nfire'
summary_peoplecaused$Year=as.factor(summary_peoplecaused$Year)

summary_peoplecaused$month=as.factor(summary_peoplecaused$month)
summary_nature$Year=as.factor(summary_nature$Year)

summary_nature$month=as.factor(summary_nature$month)
```

## Including Plots

plot the year trend

```{r , results='markup'}
library(ggplot2)

temp_plot=aggregate(nfire~Year,summary_nature,sum)

temp_plot2=aggregate(nfire~Year,summary_peoplecaused,sum)

ggplot() +geom_point(data=temp_plot, aes(x=Year, y=nfire), colour='blue') + geom_point(data=temp_plot2, aes(x=Year, y=nfire),colour='red')+labs(title='Number of Fires Each Year (red for people caused fire bule for other reasons)',y='Number of Fires')



```
plot the boxplot of the year and month to show trend

```{r, results='markup' }
library(ggpubr)
ggplot(summary_peoplecaused, mapping=aes(x=Year,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of people-caused fires for different years')+ylab('Number of Fires')

ggplot(summary_peoplecaused, mapping=aes(x=month,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of people-caused fires for different months')+ylab('Number of Fires')
ggplot(summary_nature, mapping=aes(x=Year,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of fires caused by other reasons for different years')+ylab('Number of Fires')

ggplot(summary_nature, mapping=aes(x=month,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of fires caused by other reasons for different months')+ylab('Number of Fires')

```
```{r}
ggplot(summary_peoplecaused, mapping=aes(x=Year,y=temperature)) + geom_boxplot()+ggtitle('box-plot of temperature for different years')+ylab('temperature')

ggplot(summary_peoplecaused, mapping=aes(x=month,y=temperature)) + geom_boxplot()+ggtitle('box-plot of temperature for different months')+ylab('temperature')
ggplot(summary_nature, mapping=aes(x=Year,y=soilmoisture)) + geom_boxplot()+ggtitle('box-plot of soil moisture for different years')+ylab('soil moisture')

ggplot(summary_nature, mapping=aes(x=month,y=soilmoisture)) + geom_boxplot()+ggtitle('box-plot of soil moisture for different months')+ylab('soil moisture')
ggplot(summary_nature, mapping=aes(x=Year,y=rainfall)) + geom_boxplot()+ggtitle('box-plot of rainfall for different years')+ylab('average daily rainfall')

ggplot(summary_nature, mapping=aes(x=month,y=rainfall)) + geom_boxplot()+ggtitle('box-plot of rainfall for different months')+ylab('average daily rainfall')

```

Do Anova test on the year and month statistics
```{r}
summary(aov(nfire~Year,summary_peoplecaused))
summary(aov(nfire~month,summary_peoplecaused))
summary(aov(nfire~Year,summary_nature))
summary(aov(nfire~month,summary_nature))
summary(aov(temperature~Year,summary_peoplecaused))
summary(aov(temperature~month,summary_peoplecaused))
summary(aov(soilmoisture~Year,summary_nature))
summary(aov(soilmoisture~month,summary_nature))
summary(aov(rainfall~Year,summary_nature))
summary(aov(rainfall~month,summary_nature))
```

try to make correlation check with numeric variable
```{r}
library(stringr)
temp=str_remove(summary_nature$Budget,"[$]")
temp=str_remove_all(temp,"[,]")
summary_nature$Budget=as.numeric(temp)
temp=str_remove(summary_peoplecaused$Budget,"[$]")
temp=str_remove_all(temp,"[,]")
summary_peoplecaused$Budget=as.numeric(temp)
cor_nature=cor(summary_nature[c(4:9)])

library(corrplot)
corrplot(cor_nature,method='number')
cor_people=cor(summary_peoplecaused[c(4:9)])


corrplot(cor_people,method='number')
summary_nature$total=summary_nature$n+summary_peoplecaused$n
cor_total=cor(summary_nature[c(4,5,6,8,9,10)])


corrplot(cor_total,method='number',type = 'lower', diag = TRUE)
```


create model for nature reason
and check their summary, use vif to determine the variable useage

use residual plot and qq-plot to check their normality
```{r}

model1=lm(total~temperature,data=summary_nature)
summary(model1)

model2=lm(total~temperature+soilmoisture,data=summary_nature)
plot(model2)
summary(model2)
model3=lm(total~temperature+soilmoisture+rainfall,data=summary_nature)
summary(model3)

library(car)
vif(model3)
vif(model2)
model4=lm(total~soilmoisture,data=summary_nature)

summary(model4)



model5=lm(log(total)~temperature+soilmoisture,data=summary_nature)
residualPlot(model5)
plot(model5)
summary(model5)
vif(model5)
```
