---
title: "WildFire_Project Analysis"
author: "Arundhati G., Zhongyang H., Adarsh K., Sumanth N."
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('ezids')
library(stringr)
```


```{r }
WildfireData <- read.csv('final_wildfire.csv')
summary_nature=read.csv('summary_nature.csv')
summary_peoplecaused=read.csv('summary_peoplecaused.csv')
fire_budget  <- read.csv("fire_suppression.csv")
```

```{r}

Avg_Temp <- WildfireData$tair_day_livneh_vic
Avg_SoilMoisture <- WildfireData$soilmoist1_day_livneh_vic
Avg_Rainfall <- WildfireData$rainfall_day_livneh_vic

```

##Five Point Summary 
```{r, results="asis"}
temp=str_remove(fire_budget$Budget,"[$]")
temp=str_remove_all(temp,"[,]")
fire_budget$Budget=as.numeric(temp)

temp=str_remove(WildfireData$Budget,"[$]")
temp=str_remove_all(temp,"[,]")
WildfireData$Budget=as.numeric(temp)

xkablesummary(WildfireData)

```

## HISTOGRAMS

```{r, results="markup"}

library(ggplot2)
library(gridExtra)
#Average Temperature
TempHist <- ggplot(WildfireData, aes(Avg_Temp))+ 
  geom_histogram(binwidth = 0.5, bins = 100, col="black", fill="light blue 2") +
  labs(x="Avg. Temp (C)", y="Frequency", title="HISTOGRAM: Average Temprature") 


#Average Soil Moisture
SoilHist <- ggplot(WildfireData, aes(Avg_SoilMoisture))+ 
  geom_histogram(binwidth = 0.5, bins = 100, col="black", fill="orange red 2") +
  labs(x="Avg. Soil Moisture", y="Frequency", title="HISTOGRAM: Average Soil Moisture") 

#Average Rainfall
RainHist <- ggplot(WildfireData, aes(Avg_Rainfall))+ 
  geom_histogram(binwidth = 0.5, bins = 100, col="black", fill="green 3") +
  labs(x="Avg. Rainfall", y="Frequency", title="HISTOGRAM: Average Rainfall") 

#Wildfire Count by Year
CountHist <- ggplot(WildfireData, aes(Year))+ 
  geom_histogram(binwidth = 0.10, bins = 100, col="black", fill="yellow", stat="count") +
  labs(x="Years", y="Frequency of Wildfires", title="Wildfires count by year", )

Histograms <- grid.arrange(TempHist, SoilHist, RainHist, CountHist, ncol=2, nrow=2)

ggsave("Histograms.jpg", plot = Histograms)
```


##Bar Graphs
```{r, results="markup"}

#Fire Size
FireBar <- ggplot(data = WildfireData, aes(x = FIRE_SIZE_CLASS)) +
  geom_bar(col="black", fill="orange")+
  labs(x="Fire Size Class", y="Frequency", title="Frequency of Wildfires by Size Classes") 


#Years
YearsBar <- ggplot(data = WildfireData, aes(x = Year)) +
  geom_bar(col="black", fill="yellow")+
  labs(x="Years", y="Frequency", title="Frequency of Wildfires by Year")

#Budget
BudgetBar <- ggplot(data = WildfireData, aes(x = Budget)) +
  geom_bar(col="black", fill="Pink 2")+
  labs(x="Budget", y="Frequency", title="Frequency of Wildfires by Budget")


grid.arrange(FireBar, YearsBar, nrow=2)


```

##Pie Charts
```{r, results="markup"}
lbls <- c("A", "B", "C", "D", "E", "F", "G", "E")
jpeg("sizeclasspie.jpeg")
sizeclasspie<- pie((table(WildfireData$FIRE_SIZE_CLASS)), col=rainbow(length(lbls)), main="Pie Chart of Fire Size Class")

```

```{r, results="markup"}

jpeg("cause_descrpPie.jpeg")

lbls <- c("Lightning", "Eqipment Use", "Smoking", "Campfire", "Debris Burning", "Railroad", "Arson", "Children", "Misc." )
cause_descrpPie <- pie((table(WildfireData$STAT_CAUSE_DESCR)), col=rainbow(length(lbls)), main="Pie Chart of Wildfire Cause")

```

```{r}
library(dplyr)
library(ggplot2)
yearly_count <- WildfireData %>% count(Year)
colnames(yearly_count) <- c("Year", "Count")
ggplot(yearly_count, aes(x=Year, y=Count, group=1)) + geom_line() + ggtitle("Yearly Recorded Fires")
```
```{r}
temp=str_remove(fire_budget$Budget,"[$]")
temp=str_remove_all(temp,"[,]")
fire_budget$Budget=as.numeric(temp)
ggplot(fire_budget, aes(x=Year, y=Budget, group=1)) + geom_line() + ggtitle("California Fire Suppression Budget 1979-2021")
```

```{r}
dat <- aggregate(FIRE_SIZE ~ Year, WildfireData, mean)
ggplot(dat, aes(x=Year, y=FIRE_SIZE, group=1)) + geom_line() + ggtitle("Wildfire Sizes (1992-2013)") + ylab("Fire Size")

```

## R Markdown
Hu Zhongyang part
intialize data

```{r }
final_fire=read.csv('final_wildfire.csv')
summary_nature=read.csv('summary_nature.csv')
summary_peoplecaused=read.csv('summary_peoplecaused.csv')
colnames(summary_nature)[4]='temperature'
colnames(summary_nature)[5]='soilmoisture'
colnames(summary_nature)[6]='rainfall'
colnames(summary_nature)[7]='nfire'
colnames(summary_peoplecaused)[4]='temperature'
colnames(summary_peoplecaused)[5]='soilmoisture'
colnames(summary_peoplecaused)[6]='rainfall'
colnames(summary_peoplecaused)[7]='nfire'
summary_peoplecaused$Year=as.factor(summary_peoplecaused$Year)

summary_peoplecaused$month=as.factor(summary_peoplecaused$month)
summary_nature$Year=as.factor(summary_nature$Year)

summary_nature$month=as.factor(summary_nature$month)
```

## Including Plots

plot the year trend

```{r , results='markup'}
library(ggplot2)

temp_plot=aggregate(nfire~Year,summary_nature,sum)

temp_plot2=aggregate(nfire~Year,summary_peoplecaused,sum)

ggplot() +geom_point(data=temp_plot, aes(x=Year, y=nfire), colour='blue') + geom_point(data=temp_plot2, aes(x=Year, y=nfire),colour='red')+labs(title='Number of Fires Each Year (Red for people-caused, Blue for other reasons)',y='Number of Fires')



```
plot the boxplot of the year and month to show trend

```{r, results='markup' }
library(ggpubr)
ggplot(summary_peoplecaused, mapping=aes(x=Year,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of people-caused fires for different years')+ylab('Number of Fires')

ggplot(summary_peoplecaused, mapping=aes(x=month,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of people-caused fires for different months')+ylab('Number of Fires')
ggplot(summary_nature, mapping=aes(x=Year,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of fires caused by other reasons for different years')+ylab('Number of Fires')

ggplot(summary_nature, mapping=aes(x=month,y=nfire)) + geom_boxplot()+ggtitle('box-plot of number of fires caused by other reasons for different months')+ylab('Number of Fires')

```
```{r}
ggplot(summary_peoplecaused, mapping=aes(x=Year,y=temperature)) + geom_boxplot()+ggtitle('box-plot of temperature for different years')+ylab('temperature')

ggplot(summary_peoplecaused, mapping=aes(x=month,y=temperature)) + geom_boxplot()+ggtitle('box-plot of temperature for different months')+ylab('temperature')
ggplot(summary_nature, mapping=aes(x=Year,y=soilmoisture)) + geom_boxplot()+ggtitle('box-plot of soil moisture for different years')+ylab('soil moisture')

ggplot(summary_nature, mapping=aes(x=month,y=soilmoisture)) + geom_boxplot()+ggtitle('box-plot of soil moisture for different months')+ylab('soil moisture')
ggplot(summary_nature, mapping=aes(x=Year,y=rainfall)) + geom_boxplot()+ggtitle('box-plot of rainfall for different years')+ylab('average daily rainfall')

ggplot(summary_nature, mapping=aes(x=month,y=rainfall)) + geom_boxplot()+ggtitle('box-plot of rainfall for different months')+ylab('average daily rainfall')

```

Setting up different groups for the first hypothesis test. We are looking to compare different
classes of wildfires and how different conditions may have been.
```{r}
classA <- WildfireData[WildfireData$FIRE_SIZE_CLASS == 'A',]
classB <- WildfireData[WildfireData$FIRE_SIZE_CLASS == 'B',]
classC <- WildfireData[WildfireData$FIRE_SIZE_CLASS == 'C',]
classD <- WildfireData[WildfireData$FIRE_SIZE_CLASS == 'D',]
classE <- WildfireData[WildfireData$FIRE_SIZE_CLASS == 'E',]
classF <- WildfireData[WildfireData$FIRE_SIZE_CLASS == 'F',]
classG <- WildfireData[WildfireData$FIRE_SIZE_CLASS == 'G',]
```

When comparing the conditions during the smallest wildfires to the largest wildfires,
it  appears that air temperature was lower, soil moisture was higher, and rainfall was
higher during less  intense wildfires.

```{r}
t.test(classA$tair_day_livneh_vic, classG$tair_day_livneh_vic, var.equal = TRUE)
t.test(classA$soilmoist1_day_livneh_vic, classG$soilmoist1_day_livneh_vic, var.equal = TRUE)
t.test(classA$rainfall_day_livneh_vic, classG$rainfall_day_livneh_vic, var.equal = TRUE)
```
```{r}
code1 <- WildfireData[WildfireData$STAT_CAUSE_CODE == 1,]
code2 <- WildfireData[WildfireData$STAT_CAUSE_CODE == 2,]
code7 <- WildfireData[WildfireData$STAT_CAUSE_CODE == 7,]
```
Now,  we will compare different groups of wildfires- categorized by their causes

Code 1: Lightning
Code 2: Equipment Use
Code 7: Arson

When looking  at  the wildfires caused by  Lightning versus those caused by Equipment
Use,  average temperature, soil moisture, and rainfall in CA were significantly  different.
In particular, during Lighting-caused wildfires, air temperature  was higher, soil moisture
was lower, and rainfall was higher.
```{r}
t.test(code1$tair_day_livneh_vic, code2$tair_day_livneh_vic, var.equal=TRUE)
t.test(code1$soilmoist1_day_livneh_vic, code2$soilmoist1_day_livneh_vic, var.equal=TRUE)
t.test(code1$rainfall_day_livneh_vic, code2$rainfall_day_livneh_vic, var.equal=TRUE)
```

When comparing the wildfires caused by Lightning versus those caused by Arson,
it appears that the air temperature, soil moisture, and average rainfall in CA
were significantly different. In  particular, during lightning-caused wildfires,
air temperature was higher, soil moisture was lower, and average rainfall
was higher.
```{r}
t.test(code1$tair_day_livneh_vic, code7$tair_day_livneh_vic, var.equal=TRUE)
t.test(code1$soilmoist1_day_livneh_vic, code7$soilmoist1_day_livneh_vic, var.equal=TRUE)
t.test(code1$rainfall_day_livneh_vic, code7$rainfall_day_livneh_vic, var.equal=TRUE)
```
When looking at conditions during arson-caused wildfires versus equipment use-caused
wildfires, it appears that air  temperature and soil moisture were significantly  different.
In  particular, during arson-caused fires, air temperature was lower and soil moisture was
higher.
```{r}
t.test(code7$tair_day_livneh_vic, code2$tair_day_livneh_vic, var.equal=TRUE)
t.test(code7$soilmoist1_day_livneh_vic, code2$soilmoist1_day_livneh_vic, var.equal=TRUE)
t.test(code7$rainfall_day_livneh_vic, code2$rainfall_day_livneh_vic, var.equal=TRUE)

```

Do Anova test on the year and month statistics
```{r}
summary(aov(nfire~Year,summary_peoplecaused))
summary(aov(nfire~month,summary_peoplecaused))
summary(aov(nfire~Year,summary_nature))
summary(aov(nfire~month,summary_nature))
summary(aov(temperature~Year,summary_peoplecaused))
summary(aov(temperature~month,summary_peoplecaused))
summary(aov(soilmoisture~Year,summary_nature))
summary(aov(soilmoisture~month,summary_nature))
summary(aov(rainfall~Year,summary_nature))
summary(aov(rainfall~month,summary_nature))
```

try to make correlation check with numeric variable
```{r}
temp=str_remove(summary_nature$Budget,"[$]")
temp=str_remove_all(temp,"[,]")
summary_nature$Budget=as.numeric(temp)
temp=str_remove(summary_peoplecaused$Budget,"[$]")
temp=str_remove_all(temp,"[,]")
summary_peoplecaused$Budget=as.numeric(temp)
cor_nature=cor(summary_nature[c(4:9)])

library(corrplot)
corrplot(cor_nature,method='number')
cor_people=cor(summary_peoplecaused[c(4:9)])


corrplot(cor_people,method='number')
summary_nature$total=summary_nature$n+summary_peoplecaused$n
cor_total=cor(summary_nature[c(4,5,6,8,9,10)])


corrplot(cor_total,method='number',type = 'lower', diag = TRUE)
```


create model for nature reason
and check their summary, use vif to determine the variable useage

use residual plot and qq-plot to check their normality
```{r}

model1=lm(total~temperature,data=summary_nature)
summary(model1)

model2=lm(total~temperature+soilmoisture,data=summary_nature)
plot(model2)
summary(model2)
model3=lm(total~temperature+soilmoisture+rainfall,data=summary_nature)
summary(model3)

library(car)
vif(model3)
vif(model2)
model4=lm(total~soilmoisture,data=summary_nature)

summary(model4)



model5=lm(log(total)~temperature+soilmoisture,data=summary_nature)
residualPlot(model5)
plot(model5)
summary(model5)
vif(model5)
```
